Type Evolution in Programming Systems

### 2024-05-25 paper.tex
move related work to end

\begin{comment}

@@

\begin{comment}
\section{Rename field; Add new required field}
Starter problems suggested by Geoffrey. These are already directly supported in SQL, so there is good support in SQL schema change tools, but they don't address how to synchronize the change with live code refactoring or local-first code deployment.
\end{comment}

\begin{comment}
\section{Cambria (Geoffrey/pvh)}
\citet{Cambria} exhumed and studied the old problem of schema change, primarily in the context of local-first software. In this paper we add more scenarios and the complementary problems of live programming. The primary goal of local-first software is user autonomy, which could include letting users stay on older versions of application code while continuing to collaborate with newer versions. This problem is a nice counterpart to that of Divergence Control. They both require a kind of bidirectional schema change, with Cambria building upon research on bidirectional programming with lenses, while Divergence Control is more closely connected to version control techniques. Appendix III of \cite{Cambria} discusses the tricky problem of converting between scalars and arrays (which in data modeling is called a multiplicity change). That could be the challenge problem right there.

The Cambria paper also presents motivating examples of schema change problems in Stripe, Kafka, and Mastodon. Perhaps one of those examples could be crafted into a challenge problem.
\end{comment}

\begin{comment}
\section{Denormalization (Matt Tantaman)}
Matt points out Extract Entity is a normalization so we ought to complement it with a denormalization. Note that copying/duplication is a technical problem in some formalisms like lenses. Denormalization is often done for performance reasons but that isn't an issue in our small-scale scenarios. What is a small-scale use case for denormalization?

There are good cases for denormalizing from 1NF, i.e. nesting structures. Local-first software seems focused on JSON data rather than relational. An extension of Extract Entity is \textit{Nest Entity} where a table of orders are nested inside each customer.
\end{comment}

>>>>

@@

\begin{comment}
From Tijs:

BX community:
https://www.cs.cornell.edu/~jnfoster/papers/grace-report.pdf
(overview paper)

Isil Dilig's stuff;
https://www.cs.utexas.edu/~isil/pldi19b.pdf
https://www.cs.utexas.edu/~isil/vldb20.pdf

Eelco Visser/Sander Vermolen
https://dblp.org/pid/10/6098.html
(we’ve discussed the catalog before)

On the code-database co-evolution problem:
https://link.springer.com/chapter/10.1007/11877028_17
(little less technical than Isil’s stuff)

Coupled transformations:
https://dl.acm.org/doi/10.1007/978-3-540-69611-7_19
https://www.sciencedirect.com/science/article/pii/S1571066108003058
https://link.springer.com/chapter/10.1007/978-3-540-87875-9_45

(from the MDD angle, only the tip of the ice berg…)

Divergence in model-driven:
https://dl.acm.org/doi/10.1145/2000410.2000414

The seminal model-diffing paper:
http://btn1x4.inf.uni-bayreuth.de/publications/dotor_buchmann/SCM/ChefRepo/DiffUndMerge/Alanen2003.pdf
(Diffing, versioning, collaborative editing, it’s all well researched in MDD)
